---
title: "A Provocation: Visualizing Modeled Social Networks"
author: "jimi adams"
date: "`r format(Sys.Date(), '%Y %B %d')`"
output:
  pdf_document: default
  html_document: default
bibliography: modviz.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Abstract**: Network scholarship regularly leverages visualizations. They are thought to enhance readers' recall, interpretation, and simplification of scholar's intended takeaways. However, by representing the specifics of single empirical cases, these visualization strategies may not optimize how well they accomplish our scientific purposes. In particular, they may: (1) be overly complex in ways that obfuscate what we can learn from them, (2) draw attention to the particularities of the case, rather than the general principles the case is examined to characterize, (3) be difficult to precisely replicate, raising questions about consistency in interpretation, and (4) introduce potential risks that could violate research ethical standards. To address these, I provoke the field to consider whether presenting *modeled* network results--in lieu of empirical cases--retains the aims of network visualization, while potentially overcoming some of the limitations of existing practice.

\textcolor{red}{NOTE: This is very much a fledgling idea that isn't yet ready for citation/sharing. If you're interested in helping me bring this to life (I'm presently envisioning submitting as a Viz paper to *Socius* - but it seems to be rapidly evolving beyond something that would fit there. I have no idea where it *does* fit), please reach out. Otherwise, for now, think of it as a musing...}

Social network research relies heavily on data visualization [@freeman_development_2004]. These visualizations can serve to characterize, simplify, and focus theoretical or empirical claims [@moody_dynamic_2005; @rawlings_how_2023]. They are often more readily digestible than tables of seemingly endless coefficients, significance tests, and summary statistics [@healy_data_2019; @schwabish_better_2021]. Additionally, with the increasing ubiquity of network verbiage in society, they can even convey ideas in ways that require less "translation" work for audiences who sit outside of academic research niches [@sayama_what_2016; @harrington_commentary_2013]. Quite simply, network visualization is not only common, but can be incredibly useful [@NetCanvas].^[As with any data presentation strategy--visual or otherwise--there are also ample examples of how to do this poorly (e.g., what have come to be known as unintelligible network "hairballs"); I have a few published visuals of my own I'd *substantially* rework if given the opportunity today.]

Traditionally, this has taken the form of directly visualizing instances(s) of empirical social networks--whether snapshots of single data waves, or evolving network "movies" for dynamic data structures [@katyao_2025].^[These typically occur *after* data pre-processing has taken place (e.g., cleaning, recoding, etc.), but no further data reduction steps (i.e., are not the product of data summaries or analytic interpretations).] While this practice predominates, it's worth asking whether alternatives may better serve the *purposes* of network visualization.

\textcolor{red}{I'm working here to re-frame this around what's proposed, and highlight why modeled results do better, rather than first centering a single problem (e.g., ethical considerations - even though that's where this idea intiially sprung from for me). In part, this shift is motivated by suggestions (e.g., from Jim) that some may not view all the potential constraints that led me here as being similarly problematic. So I might need to motivate/sell it more broadly (If I buy it in those terms).}

\textcolor{red}{I think before getting into the potential constraints of the existing approach, it may be helpful here to elaborate more what they do **well**.}

1. **Clarity** If visualizations' aim to communicate scientific ideas, modeled network visualizations may do this better than empirical network graphs. When I teach introductory science of science principles, I vociferously attempt to convince students that science's primary product is a collection of theories that are more-or-less empirically supported, not *merely* that collection of facts/findings [@zimring2019]. Empirical results flesh out what theories mean, where they're supported, and where they need further refinement [@kuhn1962]; they don't "stand on their own" [@Weber1949]. If network representations convey single empirical cases, we risk drawing attention to the particularities of the example, and distracting from the broader principles that case is meant to illustrate--limiting how they produce understanding among readers of what we have learned *theoretically* from empirical analyses. Visualizing modeled results can focus our (readers') attention on what we're really claiming to have learned.

1. **Simplification** Theoretical models, *by necessity*, are simplifications of complex and messy real world processes. 

1. **Systematization**

1. **Safety** Another key potential limitation that arise form network visualizations is that they can introduce unique ethical challenges to presenting research [@adams_gathering_2019]. Especially in "closed population" settings [but not only there - c.f., @zimmer_but_2010], the potential for deductive disclosure is broad in any presentation of multiple variables [@arfer_american_2019], and network data are uniquely primed to allow serial-revelations of participants' identities---i.e., the ability to identify one node in a graph (e.g., myself) often heightens the capacity of identifying others' identities in turn (e.g., my friends). To this end, our frequent practice of plotting empirical networks from data collection efforts \textcolor{red}{(do I want to tell my story of Add Health in my 4-cylce viz paper here?)} may undermine principles of confidentiality and minimal risk in ways that should lead us to consider alternate visualization strategies that can simultaneously leverage these strengths while minimizing these potential risks [@ethicswhite_2026].

## Plot Modeled, not Raw, Networks

My fundamental proposition is simple - instead of plotting the "raw" empirical networks directly presenting data that has been collected and researchers in-turn analyze, we can produce visualizations of modeled network results that arise from those analyses. This is not a ground-breaking suggestion. In fact, it's a practice many statistical modelers of social network data already enact in practice. For example,  "eye-balling" consistency of modeled results with observed networks--in tandem with formal statistical tests for goodness of fit--is even typically taught as a helpful practice in model development and fitting processes [@krivitsky_ergm_2023]. However, while we use these heuristic network visualizations to improve model fit (whether for capturing theoretical limitations, or for enhancing stability of estimates), when presenting networks in publications, authors still often resort to presenting the "real" network - often as a first-pass descriptive task to characterize the nature of the network(s) that the subsequent analyses are intended to explain. I've done this multiple times myself.

But if our aims in visualizing networks are to clearly, simply, systematically, and ethically convey scientifically useful network visualizations, there's no reason our visualizations need to be of the raw data. Modeled network representations can better serve our purposes across these aims.

Let me draw a parallel here. Logistic regression coefficients (just like exponential random graph models) are notoriously hard to make substantive sense of on their own (e.g., significance and magnitude of effects are frequently conflated in results interpretations by writers and readers alike). A common solution to this has been for researchers to (1) gather, pre-process, then provide a descriptive summary of their data to given an overview sense of what the sample being analyzed actually represents, and how key focal features are distributed in the population; (2) conduct statistical analyses to investigate the researchers' theoretical aims, which are summarized with traditional combinations of betas, standard errors, log-likelihoods, etc. to represent the fit of those claims to empirical estimation, then (3) transform those modeled results into "predicted probabilities" in ways that help interpret what those modeled results "look like" in their implications [@fox_effect_2006]. To "tell the story" that arises from the model results we don't return to the raw data, provide the individual-level combination of variables for some subset of the population and rely on those case's patterns to convey the implications of our research [in the rare cases that people do, they are often "synthetic" cases, not real ones, e.g.,  @ayala_image_2019]. To do so would not only fail to clearly tell the aggregate story we've learned in the course of our theorization and analysis; it would also violate a number of human subjects protections researchers are obligated to follow. So why in the network world would we need to effectively present the "real" pattern of relationships from our data as a means to interpret and convey our findings in visual form?

## Implementation

Instead, I suggest that we can follow a similar logic to develop presentation- and publication-worthy network visualizations from our modeled results in ways that accomplish the aims we have for visualizations--perhaps even more clearly than the directly observed versions--while maintaining protections against potential ethical gaps (e.g., deductive disclosure). The process I'm recommending follows the same logic as for regression-type models laid out above:

1. Present descriptives as *summary* statistics of the characteristics of one's network(s) being investigated [@Neal_2024].
1. Conduct whatever form of analyses are appropriate to the theoretical aims of the research question(s).
1. Compile appropriate analytic overviews of results (e.g., statistics) into summary form.
1. Visualize "predicted" network(s) that are specified by the corresponding network analytic (model) results from which the researchers' conclusions are reached.

## Still to be filled in - Example

\textcolor{red}{Work out the "aims" of visualization a bit more concretely, and elaborate how modeled viz may do this even better than raw data viz can.}

\textcolor{red}{Let's illustrate this with an example.}

```{r ergm}
```

# References