---
title: "Backbone Illustration"
author: "jimi adams"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 8,
  dpi = 300
)
```

# Overview

This document builds a visualization of the rap samples that Jenn Lena used in her dissertation and subsequent *Poetics* paper -@lena2004. Mark Pachucki provided the raw files that I am working from as part of a replication package for their related co-authored later *Poetics* paper -@lena2013.


## Setup 

Just packages, data, and re-labeling to get things started.

```{r packages}
library(dplyr)
library(tidyr)
library(igraph)
library(ggraph)
library(ggplot2)
library(scales)
library(purrr)
library(tibble)
library(rlang)
library(backbone)
library(patchwork)

df <- readRDS("data/JL_samples.rds")
```

Just to clarify what's in that file:
| var name | brief derscription |
|---|---|
| `src_id` | source/sample song ID |
| `src_gen` | source/sample genre |
| `src_art` | source/sample artist label |
| `src_sng` | source/sample song title |
| `tar_id` | target/rap song ID |
| `tar_art` | target/rap artist label |
| `tar_sng` | target/rap song title |
| `yr` | year |
| `art_tar` | target/rap artist ID |
| `art_src` | source/sample artist ID |

We'll convert those to *slightly* more informative variable names; typing efficiency be damned, mental space is too precious.

```{r standardize-data, eval=T}
samp <- df %>%
  transmute(
## sampling details
    rap_song_id = tar_id,
    rap_artist_id = art_tar,
    rap_artist = tar_art,
    rap_song = tar_sng,
    year = as.integer(yr),
## sampled details
    sampled_song_id = src_id,
    sampled_artist_id = art_src,
    sampled_artist = src_art,
    sampled_song = src_sng,
    sampled_genre = as.character(src_gen)
  ) 
```

## The "raw" samples network

We're going to first build a bipartite incidence matrix of:

- **rap (sampling) **songs** on the rows** by
- **sampled *artists* on the columns**

Note these are weighted, because there are times where the same song multiply samples the same artist (in fact several 3 times!).

```{r build-incidence}

##  first as a graph
g <- graph_from_data_frame(
  data.frame(from = samp$rap_song_id, to = samp$sampled_artist_id), directed = FALSE)

## converting to 2-mode
V(g)$type <- V(g)$name %in% samp$rap_song_id

## converting to 2-mode
V(g)$type <- V(g)$name %in% samp$rap_song_id
## collapsing multiple ties to edge weights
E(g)$weight <- 1 #needs a base
gw <- igraph::simplify(g, remove.multiple = TRUE, remove.loops = TRUE,
  edge.attr.comb = list(weight = "sum", .default = "ignore"))


## adding year as an *edge* attribute
edge_lookup <- samp %>%
  distinct(rap_song_id, sampled_artist_id, .keep_all = TRUE) %>%
  select(rap_song_id, sampled_artist_id, year) %>%
  # igraph stores as character, so making merge-able
  mutate(
    rap_song_id = as.character(rap_song_id),
    sampled_artist_id = as.character(sampled_artist_id)
  )
edges_joined <- igraph::as_data_frame(gw, what = "edges") %>%
  left_join(edge_lookup, by = c("from" = "rap_song_id", "to" = "sampled_artist_id"))
E(gw)$year <- edges_joined$year


## let's reattach the nodal attributes for potential use later
sampled_artist_attributes <- samp %>%
  select(sampled_artist_id, sampled_artist, sampled_genre) %>%
  distinct(sampled_artist_id, .keep_all = TRUE)
# I'm not sure how much we're going to use the sampling side here, but just in case, those too.
rap_song_attributes <- samp %>%
  select(rap_song_id, rap_artist, rap_song, year) %>%
  distinct(rap_song_id, .keep_all = TRUE)

V(gw)$artist <- ifelse(
  V(gw)$type, 
  rap_song_attributes$rap_artist[match(V(gw)$name, rap_song_attributes$rap_song_id)],
  sampled_artist_attributes$sampled_artist[match(V(gw)$name, sampled_artist_attributes$sampled_artist_id)]
)

# I can't imagine I'm going to use this in any plots, but it doesn't hurt to carry it.
V(gw)$song_title <- ifelse(
  V(gw)$type,
  rap_song_attributes$rap_song[match(V(gw)$name, rap_song_attributes$rap_song_id)],
  NA # doesn't apply to sampled artists
)
# Duplicate of edge attribute above, but I'm not positive which way I'll use it.
V(gw)$year <- ifelse(
  V(gw)$type,
  rap_song_attributes$year[match(V(gw)$name, rap_song_attributes$rap_song_id)],
  NA # doesn't apply to sampled artists
)

# This one's key for the analyses here
V(gw)$genre <- ifelse(
  !V(gw)$type,
  sampled_artist_attributes$sampled_genre[match(V(gw)$name, sampled_artist_attributes$sampled_artist_id)],
  NA # I don't want to have a genre attached to the sampling songs
)
## making it plotable
unique_genres <- unique(na.omit(V(gw)$genre))
genre_colors  <- rainbow(length(unique_genres))
names(genre_colors) <- unique_genres
# sampling songs (type = TRUE) become white; sampled artists color by genre
V(gw)$color <- ifelse(
  V(gw)$type, 
  "white", 
  genre_colors[V(gw)$genre]
)

### Given a bunch of data manipulations above, I want to run one sanity check on something that was multiply assigned and thus "checkable" to ensure all was behaving as anticipated:
# dq_check <- igraph::as_data_frame(gw, what = "edges") %>%
#   mutate(node_year = V(gw)$year[match(from, V(gw)$name)])
# any(dq_check$year != dq_check$node_year) # is FALSE (as should be)

## and converting that to matrix form (for the backbone)
mw <- t(as.matrix(as_biadjacency_matrix(gw, attr = "weight")))

```

# Generating an *Exploratory* plot
I'm actually going to come back to this one last, ironically enough to the point of this exercise.

```{r plot-panelA, include=T}
# We're going to add sampling degree to the sampled artists as a nodal attribute:
artist_str <- strength(gw, v = V(gw)[!V(gw)$type])
vsize <- rep(2, vcount(gw)) # fixing size for sampling nodes
vsize[!V(gw)$type] <- scales::rescale(log1p(artist_str), to = c(1, 4))
names(vsize) <- V(gw)$name
V(gw)$vsize <- vsize

source("src/igraph-triangles.R")
V(gw)$shape <- ifelse(V(gw)$type, "triangle", "circle")

# likely only going to plot the giant component, but let's carry the small ones just in case
components_list <- decompose(gw)
components_list <- components_list[order(sapply(components_list, vcount), decreasing = TRUE)]
g_core <- components_list[[1]]
layout_core <- layout_with_fr(g_core)

# now we're going to strip off, and plot a grid of the smaller components
minor_components <- components_list[2:length(components_list)]
g_small <- disjoint_union(minor_components)
vsize_small <- vsize[V(g_small)$name] # just the ones that are still here
all_coords_small <- matrix(0, nrow = vcount(g_small), ncol = 2)
rownames(all_coords_small) <- V(g_small)$name
max_row_width <- 36  # The virtual horizontal canvas size for the tile panel
padding <- 1.5       # Gutter separation buffer between adjacent components
  current_x <- 0
  current_y <- 0
  max_height_in_row <- 0
  
  # Step through each minor component to translate its true scale geometry
  for (i in seq_along(minor_components)) {
    comp <- minor_components[[i]]
    comp_coords <- layout_with_fr(comp)
    
    # 1. Measure the native, honest spatial footprint of the layout
    comp_width <- max(comp_coords[, 1]) - min(comp_coords[, 1])
    comp_height <- max(comp_coords[, 2]) - min(comp_coords[, 2])
    
    # 2. Zero-center coordinates so translations don't distort shapes
    comp_coords[, 1] <- comp_coords[, 1] - mean(comp_coords[, 1])
    comp_coords[, 2] <- comp_coords[, 2] - mean(comp_coords[, 2])
    
    # 3. FIX: Determine if adding this component spills past our maximum row width boundary.
    # If it exceeds the boundary, automatically trigger a clean line break.
    if (current_x + comp_width > max_row_width && current_x > 0) {
      current_x <- 0
      current_y <- current_y - max_height_in_row - padding
      max_height_in_row <- 0
    }
    
    # 4. Map coordinates precisely onto the current visual baseline tracking point
    comp_coords[, 1] <- comp_coords[, 1] + current_x + (comp_width / 2)
    comp_coords[, 2] <- comp_coords[, 2] + current_y - (comp_height / 2)
    
    # Commit coordinates to master layout tracker matrix
    all_coords_small[V(comp)$name, ] <- comp_coords
    
    # Record the tallest component processed in this row to compute the next row offset
    if (comp_height > max_height_in_row) {
      max_height_in_row = comp_height
    }
    
    # 5. Move horizontal cursor ahead by the actual width of this component plus padding
    current_x <- current_x + comp_width + padding
  }

# exploratory plot of just the giant component
source("src/igraph-triangles.R")

# Exploratory plot of just the giant component
plot(g_core, 
     layout = layout_core, 
     vertex.size = V(g_core)$vsize, 
     vertex.label = NA,
     vertex.shape = c("circle", "triangle")[V(g_core)$type + 1], 
     margin = 0.15,
     edge.width = 0.5,
     edge.color = rgb(0.3, 0.3, 0.3, alpha = 0.15))

title(main = "Bipartite Sampling Graph", line = 2.5, cex.main = 1.5)
title(main = "large component only", line = 1.2, cex.main = 1.0, font.main = 1)
legend(
  "bottomright", 
  legend = c("node size proportional to strength", "(sampled artist) node colored by genre"),
  bty = "n",       # Removes the box border around the text
  cex = 0.8        # Scales text size (optional, 0.8 makes it slightly smaller)
)
par(mfrow = c(1, 2), mar = c(1, 1, 3, 1))
plot(g_core, 
     layout = layout_core, 
     vertex.size = V(g_core)$vsize, 
     vertex.label = NA,
     margin = 0.15,
     edge.width = 0.5,
     edge.color = rgb(0.3, 0.3, 0.3, alpha = 0.15))

plot(g_small, 
       layout = all_coords_small, 
       vertex.size = vsize_small, 
       vertex.label = NA,
       edge.width = 0.6,
       edge.color = rgb(0.3, 0.3, 0.3, alpha = 0.3))

mtext("Bipartite Rap Samples Network", 
      side = 3, 
      outer = TRUE, 
      line = 1, 
      cex = 1.5, 
      font = 2)

par(mfrow = c(1, 1))
```

I'm way down a rabbit hole or rethinking how to model Jenn's claims. Backbone appears to have been a distraction.

# Generating the "hybrid" plot (likely shoving this one in an appendix, but for a key claim).

# Generating the "explanatory" plot
```{r panelC}
V(gw)$str <- strength(gw)
table(V(gw)$str[as.numeric(V(gw)$name)<500]) # looking at the distribution (for sampled artists only)
prop <- table(V(gw)$str[as.numeric(V(gw)$name)<500])/sum(table(V(gw)$str[as.numeric(V(gw)$name)<500])) # as proportion
cdf <- cumsum(prop)
cdf95 <- 8
core <- V(gw)$name[which(V(gw)$str>cdf95 & as.numeric(V(gw)$name)<500)] # finding top 5%
# cbind( V(gw)$artist[V(gw)$name %in% core],  V(gw)$genre[V(gw)$name %in% core]) # just curious
# just checking that they're all in the gc
# core %in% V(components_list[[1]])$name

core_vs <- V(gw)[name %in% as.character(core)]
nbr_names <- V(gw)$name[unique(unlist(neighborhood(gw, order = 1, nodes = core_vs), use.names = FALSE))]
keep_vs <- V(gw)[name %in% unique(c(as.character(core), nbr_names))]
g_core_neighborhood <- induced_subgraph(gw, vids = keep_vs)

table(V(g_core_neighborhood)$type)
table(V(gw)$type)
# over 40% of the samples come from those 26 (top 5% of) sampled artists.


```