---
title: "Backbone Illustration"
author: "jimi adams"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 8,
  dpi = 300
)
```

# Overview

This document builds a visualization of the rap samples that Jenn Lena used in her dissertation and subsequent *Poetics* paper -@lena2004. Mark Pachucki provided the raw files that I am working from as part of a replication package for their related co-authored later *Poetics* paper -@lena2013.


## Setup 

Just packages, data, and re-labeling to get things started.

```{r packages}
library(dplyr)
library(tidyr)
library(igraph)
library(ggraph)
library(ggplot2)
library(scales)
library(purrr)
library(tibble)
library(rlang)
library(backbone)

df <- readRDS("data/JL_samples.rds")
```

Just to clarify what's in that file:
| var name | brief derscription |
|---|---|
| `src_id` | source/sample song ID |
| `src_gen` | source/sample genre |
| `src_art` | source/sample artist label |
| `src_sng` | source/sample song title |
| `tar_id` | target/rap song ID |
| `tar_art` | target/rap artist label |
| `tar_sng` | target/rap song title |
| `yr` | year |
| `art_tar` | target/rap artist ID |
| `art_src` | source/sample artist ID |

We'll convert those to *slightly* more informative variable names; typing efficiency be damned, mental space is too precious.

```{r standardize-data, eval=T}
samp <- df %>%
  transmute(
## sampling details
    rap_song_id = tar_id,
    rap_artist_id = art_tar,
    rap_artist = tar_art,
    rap_song = tar_sng,
    year = as.integer(yr),
## sampled details
    sampled_song_id = src_id,
    sampled_artist_id = art_src,
    sampled_artist = src_art,
    sampled_song = src_sng,
    sampled_genre = as.character(src_gen)
  ) 
```

## The "raw" samples network

We're going to first build a bipartite incidence matrix of:

- **rap (sampling) **songs** on the rows** by
- **sampled *artists* on the columns**

Note these are weighted, because there are times where the same song multiply samples the same artist (in fact several 3 times!).

```{r build-incidence}

##  first as a graph
g <- graph_from_data_frame(
  data.frame(from = samp$rap_song_id, to = samp$sampled_artist_id), directed = FALSE)

## converting to 2-mode
V(g)$type <- V(g)$name %in% samp$rap_song_id
## collapsing multiple ties to edge weights
E(g)$weight <- 1 #needs a base
gw <- igraph::simplify(g, remove.multiple = TRUE, remove.loops = TRUE,
  edge.attr.comb = list(weight = "sum", .default = "ignore"))
## and converting that to matrix form (for the backbone)
mw <- as.matrix(as_biadjacency_matrix(gw, attr = "weight"))

```

# Backbone
Let's pull off the **sampled artists'** backbone.
```{r backbone}
w <- mw %*% t(mw)
bb<- backbone(w)
```

# Generating an *Exploratory* plot
I'm actually going to come back to this one last, ironically enough to the point of this exercise.

```{r plot-panelA}
layout_full <- layout = layout_with_kk(g) # will decide later what the plot algo is

gc <- subgraph(g, clusters(g)$membership==1)
layout_gc <- layout = layout_with_kk(gc) # will decide later what the plot algo is

source("src/igraph-triangles.R")

# We're going to add sampling degree to the sampled artists as a nodal attribute:
artist_deg <- degree(g, v = V(g)[!V(g)$type])
vsize <- rep(3, vcount(g))
vsize[!V(g)$type] <- scales::rescale(log1p(artist_deg), to = c(3, 10))

# For full network
plot.igraph(g, vertex.size=3, vertex.label=NA, vertex.shape = c("circle", "triangle")[V(g)$type + 1], vertex.color= c("black", "white")[V(g)$type + 1],)

# For giant component only
plot.igraph(gc, vertex.size=3, vertex.label=NA, vertex.shape = c("circle", "triangle")[V(gc)$type + 1], vertex.color= c("black", "white")[V(gc)$type + 1],)

```

# Generating a "hybrid" plot



